Scaling up the hyperparameter grid for Naomi

### 2023-03-20 src/dev_scale-grid/scale-grid.Rmd
Check that PCA works with smaller example

---
title: "Scaling up the hyperparameter grid for Naomi"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background** We have previously fit empirical Bayes style approaches to the Naomi model, fixing hyperparameter values.

  **Task** We would like to begin to integrate over the hyperparameters using some kind of grid structure. The problem is that there are too many to use anything dense.
---

@@

```{r results="hide"}
area_merged <- read_sf(system.file("extdata/demo_areas.geojson", package = "naomi"))
pop_agesex <- read_csv(system.file("extdata/demo_population_agesex.csv", package = "naomi"))
survey_hiv_indicators <- read_csv(system.file("extdata/demo_survey_hiv_indicators.csv", package = "naomi"))
art_number <- read_csv(system.file("extdata/demo_art_number.csv", package = "naomi"))
anc_testing <- read_csv(system.file("extdata/demo_anc_testing.csv", package = "naomi"))
pjnz <- system.file("extdata/demo_mwi2019.PJNZ", package = "naomi")
spec <- naomi::extract_pjnz_naomi(pjnz)

@@

Add on the modes:

```{r}
for(j in 1:d) b_nodes[, j] <- b_nodes[, j] + m[j]
```

Adjust the weights:

```{r}
gg_s$weights
det(Lambda)
det(Lambda_s)
```

```{r}
#' @param m Mean vector
#' @param C Covariance matrix
#' @param s Small grid dimension
#' @param k Number of points per small grid dimension
pca_rescale_grid <- function(m, C, s, k) {
  lambda <- eigenC$values
  Lambda <- diag(lambda)
  E <- eigenC$vectors
  E_s <- E[, 1:s] 
  gg_s <- create_grid(d = s, k = k)
  a_nodes <- gg_s$nodes
  b_nodes <- t(E_s %*% diag(lambda[1:s]^{-0.5}) %*% t(a_nodes))
  for(j in 1:d) b_nodes[, j] <- b_nodes[, j] + m[j]
  b_weights <- gg_s$weights * sqrt(det(Lambda))
  return(list(nodes = b_nodes, weights = b_weights))
}
```

>>>>

---
title: "Scaling up the hyperparameter grid for Naomi"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background** We have previously fit empirical Bayes style approaches to the Naomi model, fixing hyperparameter values to a single value.

  **Task** We would like to begin to integrate over the hyperparameters using some kind of grid structure. The problem is that there are >20 hyperparameters, and so using any type of dense grid is not practical. In this notebook we investigate approaches to using a smaller number of grid points to cover the space, including 1) allocating points in proporiton to posterior standard deviation 2) using principle components analysis.
---

@@

```{r results="hide", message=FALSE}
area_merged <- read_sf(system.file("extdata/demo_areas.geojson", package = "naomi"))
pop_agesex <- read_csv(system.file("extdata/demo_population_agesex.csv", package = "naomi"))
survey_hiv_indicators <- read_csv(system.file("extdata/demo_survey_hiv_indicators.csv", package = "naomi"))
art_number <- read_csv(system.file("extdata/demo_art_number.csv", package = "naomi"))
anc_testing <- read_csv(system.file("extdata/demo_anc_testing.csv", package = "naomi"))
pjnz <- system.file("extdata/demo_mwi2019.PJNZ", package = "naomi")
spec <- naomi::extract_pjnz_naomi(pjnz)

@@

Put this into a function, which also adjusts according to the mean and reweights the nodes:

```{r}
#' @param m Mean vector
#' @param C Covariance matrix
#' @param s Small grid dimension
#' @param k Number of points per small grid dimension
pca_rescale_grid <- function(m, C, s, k) {
  d <- nrow(C)
  stopifnot(d == length(m))
  eigenC <- eigen(C)
  lambda <- eigenC$values
  Lambda <- diag(lambda)
  E <- eigenC$vectors
  E_s <- E[, 1:s] 
  gg_s <- create_grid(d = s, k = k)
  a_nodes <- gg_s$nodes
  b_nodes <- t(E_s %*% diag(lambda[1:s]^{-0.5}, ncol = s) %*% t(a_nodes))
  for(j in 1:d) b_nodes[, j] <- b_nodes[, j] + m[j]
  b_weights <- gg_s$weights * sqrt(det(Lambda))
  return(list(nodes = b_nodes, weights = b_weights))
}
```

Check that this works as expected for a simple example:

```{r}
mu <- c(1, 1.5)
cov <- matrix(c(2, 1, 1, 1), ncol = 2)

obj <- function(theta) {
  mvtnorm::dmvnorm(theta, mean = mu, sigma = cov)
}

grid <- expand.grid(
  theta1 = seq(-2, 5, length.out = 700),
  theta2 = seq(-2, 5, length.out = 700)
)

ground_truth <- cbind(grid, pdf = obj(grid))

plot <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
  geom_contour(col = multi.utils::cbpalette()[1]) +
  coord_fixed(xlim = c(-2, 5), ylim = c(-2, 5), ratio = 1) +
  theme_minimal()

plot

test <- pca_rescale_grid(m = mu, C = cov, s = 1, k = 5)

plot +
  geom_point(
    data = data.frame(test$nodes),
    aes(x = X1, y = X2),
    col = multi.utils::cbpalette()[2],
    inherit.aes = FALSE
  )
```

Apply it to the Naomi grids:
